---
params:
  library: Example
  unfiltered_sce: !r scpcaTools:::sim_sce()
  filtered_sce: NULL
  processed_sce: NULL
  date: !r Sys.Date()

title: "`r glue::glue('ScPCA QC report for {params$library}')`"
author: "Childhood Cancer Data Lab"
date: "`r params$date`"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
    number_sections: false
    code_download: true
---

```{r setup, message = FALSE, echo = FALSE}
# knitr options
knitr::opts_chunk$set(
  echo = FALSE
)

library(SingleCellExperiment)
library(dplyr)
library(ggplot2)

# Set default ggplot theme
theme_set(theme_bw() + 
          theme(plot.margin = margin(rep(20, 4))))
```

```{r sce_setup}
# save some typing later
library_id <- params$library
unfiltered_sce <- params$unfiltered_sce
filtered_sce <- params$filtered_sce
processed_sce <- params$processed_sce

has_filtered <- !is.null(filtered_sce)
has_processed <- !is.null(processed_sce)

# grab sample id from filtered sce, if missing set sample ID to NA
if(is.null(metadata(filtered_sce)$sample_id)){
  sample_id <- NA
} else {
  sample_id <- metadata(filtered_sce)$sample_id
}

# if there is no filtered sce, use the unfiltered for both
if (!has_filtered){
  filtered_sce <- unfiltered_sce
}

# add cell stats if missing
if (is.null(unfiltered_sce$sum)){
  unfiltered_sce <- scuttle::addPerCellQCMetrics(unfiltered_sce)
}
if (is.null(filtered_sce$sum)){
  filtered_sce <- scuttle::addPerCellQCMetrics(filtered_sce)
}
if (is.null(filtered_sce$subsets_mito_percent)){
  filtered_sce$subsets_mito_percent <- NA_real_
  skip_miQC <- TRUE
}else{
  skip_miQC <- FALSE
}

# try to add miQC if it is missing
if (is.null(metadata(filtered_sce)$miQC_model) & !skip_miQC){
  filtered_sce <- scpcaTools::add_miQC(filtered_sce)
}

## Check for additional modalities
modalities <- c("RNA-seq")

has_cite <- "CITEseq" %in% altExpNames(filtered_sce)
if (has_cite){
  modalities <- c(modalities, "CITE-seq")
}

# check for cellhash to add to list of modalities
has_cellhash <- "cellhash" %in% altExpNames(filtered_sce)
if(has_cellhash){
  modalities <- c(modalities, "Multiplex")
}
```

# Processing Information for `r library_id`

```{r, results='asis'}
# only print out multiplexed warning if more than one sample is present
has_multiplex <- length(sample_id) > 1

if(has_multiplex){
  # convert sample ID to bullet separated list 
  multiplex_samples <- paste0("<li>", paste(sample_id, collapse = '</li><li>', "</li>"))
  glue::glue("
    <div class=\"alert alert-warning\">
    
    This library is multiplexed and contains data from more than one sample.  
    Data from the following samples are included in this library:
    
    {multiplex_samples}
    
    </div>
  ")
}

```

## Raw Sample Metrics

```{r }
# extract sce metadata containing processing information as table
unfiltered_meta <- metadata(unfiltered_sce) 

sample_information <- tibble::tibble(
  "Library id" = library_id,
  "Sample id" = paste(sample_id, collapse = ", "),
  "Tech version" = format(unfiltered_meta$tech_version), # format to keep nulls
  "Data modalities" = paste(modalities, collapse = ", "),
  "Cells reported by alevin-fry" = 
    format(unfiltered_meta$af_num_cells, big.mark = ',', scientific = FALSE),
  "Number of genes assayed" = 
    format(nrow(unfiltered_sce), big.mark = ',', scientific = FALSE),
  "RNA-seq reads sequenced" = 
    format(unfiltered_meta$total_reads, big.mark = ',', scientific = FALSE),
  "Percent RNA-seq reads mapped to transcripts" = 
    paste0(round((unfiltered_meta$mapped_reads/unfiltered_meta$total_reads) *100, 2), "%")
)
if (has_cite){
  cite_exp <- altExp(filtered_sce, "CITEseq")
  cite_meta <- metadata(cite_exp)
  
  sample_information <- sample_information |>
    mutate(
      "Number of ADTs assayed" = 
        format(nrow(cite_exp), big.mark = ',', scientific = FALSE),
      "CITE-seq reads sequenced" = 
        format(cite_meta$total_reads, big.mark = ',', scientific = FALSE),
      "Percent CITE-seq reads mapped to ADTs" = 
        paste0(round(cite_meta$mapped_reads/cite_meta$total_reads * 100, digits = 2), "%")
    )
}
if (has_cellhash){
  multiplex_exp <- altExp(filtered_sce, "cellhash")
  multiplex_meta <- metadata(multiplex_exp)
  
  sample_information <- sample_information |>
    mutate(
      "Number of HTOs assayed" = 
        format(nrow(multiplex_exp), big.mark = ',', scientific = FALSE),
      "Multiplex reads sequenced" = 
        format(multiplex_meta$total_reads, big.mark = ',', scientific = FALSE),
      "Percent of multiplex reads mapped to HTOs" = 
        paste0(round(multiplex_meta$mapped_reads/multiplex_meta$total_reads * 100, digits = 2), "%")
    )
}


sample_information <- sample_information |>
  mutate(across(.fns = ~ifelse(.x == "NULL", "N/A", .x))) |> # reformat nulls
  t()

# make table with sample information
knitr::kable(sample_information, align = 'r') |>
  kableExtra::kable_styling(bootstrap_options = "striped",
                            full_width = FALSE,
                            position = "left") |>
  kableExtra::column_spec(2, monospace = TRUE)
```

## Pre-Processing Information

```{r }
processing_info <- tibble::tibble(
  "Salmon version"       = format(unfiltered_meta$salmon_version),
  "Alevin-fry version"   = format(unfiltered_meta$alevinfry_version),
  "Transcriptome index"  = format(unfiltered_meta$reference_index),
  "Alevin-fry droplet detection"  = format(unfiltered_meta$af_permit_type),
  "Resolution"           = format(unfiltered_meta$af_resolution), 
  "Transcripts included" = dplyr::case_when(
      format(unfiltered_meta$transcript_type) == "spliced" ~ "Spliced only",
      format(unfiltered_meta$transcript_type) == "unspliced" ~ "Spliced and unspliced",
      TRUE ~ format(unfiltered_meta$transcript_type))
  ) |>
  mutate(across(.fns = ~ifelse(.x == "NULL", "N/A", .x))) |>
  t()


# make table with processing information
knitr::kable(processing_info, align = 'r') |>
  kableExtra::kable_styling(bootstrap_options = "striped",
                            full_width = FALSE,
                            position = "left") |>
  kableExtra::column_spec(2, monospace = TRUE)
```

# RNA-seq Experiment Summary 

## Cell Statistics 

```{r}
basic_statistics <- tibble::tibble(
  "Method used to filter empty droplets"          = metadata(filtered_sce)$filtering_method, 
  "Number of cells post filtering empty droplets" = format(ncol(filtered_sce), big.mark = ","),
  "Percent of reads in cells"                     = paste0(round((sum(filtered_sce$sum)/sum(unfiltered_sce$sum))*100,2),"%"),
  "Median UMI count per cell"                     = format(median(filtered_sce$sum), big.mark = ","),
  "Median genes detected per cell"                = format(median(filtered_sce$detected), big.mark = ","),
  "Median percent reads mitochondrial"            = paste0(round(median(filtered_sce$subsets_mito_percent), 2), "%")
  )

# if processed sce exists add filtering and normalization table
if(has_processed){
  processed_meta <- metadata(processed_sce)
  
  basic_statistics <- basic_statistics |>
    mutate(
      "Method used to filter low quality cells" = format(processed_meta$ccdl_filter_method),
      "Cells after filtering low quality cells" = format(dim(processed_sce)[2], big.mark = ',', scientific = FALSE),
      "Normalization method"                    = format(processed_meta$normalization),
      "Minimum genes per cell cutoff"           = format(processed_meta$min_gene_cutoff)
    
    )
  if(processed_meta$ccdl_filter_method == "miQC") {
    basic_statistics <- basic_statistics |>
      mutate(
        "Probability of compromised cell cutoff" = format(processed_meta$prob_compromised_cutoff, big.mark = ",", scientific = FALSE)
      )
  }
}


basic_statistics <- basic_statistics |>
  mutate(across(.fns = ~ifelse(.x == "NULL", "N/A", .x))) |> # reformat nulls
  t()

# make table with basic statistics
knitr::kable(basic_statistics, align = 'r') |>
  kableExtra::kable_styling(bootstrap_options = "striped",
                            full_width = FALSE,
                            position = "left") |>
  kableExtra::column_spec(2, monospace = TRUE)

```

```{r, results='asis'}
if(metadata(filtered_sce)$filtering_method == "UMI cutoff"){
  glue::glue("
    <div class=\"alert alert-warning\">
    
    This library may contain a low number of cells and was unable to be filtered using `DropletUtils`. 
    Droplets with a total UMI count ≥ {metadata(filtered_sce)$umi_cutoff} are included in the filtered `SingleCellExperiment` object.
    
    </div>
  ")
}
```


## Knee Plot

```{r, fig.alt="Smoothed knee plot of filtered and unfiltered droplets"}
unfiltered_celldata <- data.frame(colData(unfiltered_sce)) |>
  mutate(
    rank = rank(- unfiltered_sce$sum, ties.method = "first"), # using full spec for clarity 
    filter_pass = colnames(unfiltered_sce) %in% colnames(filtered_sce)
  ) |>
  select(sum, rank, filter_pass) |>
  filter(sum > 0) # remove zeros for plotting


grouped_celldata <- unfiltered_celldata |>
  mutate(rank_group = floor(rank / 100) )|>
  group_by(rank_group) |>
  summarize(
    med_sum = median(sum),
    med_rank = median(rank),
    pct_passed = sum(filter_pass) / n() * 100
  ) 

top_celldata <- unfiltered_celldata |>
  filter(rank <= 50) |>
  mutate(filter_pct = ifelse(filter_pass, 100, 0))

ggplot(grouped_celldata, aes(x = med_rank, y = med_sum, color = pct_passed)) + 
  geom_point(mapping = aes(x = rank, y = sum, color = filter_pct),
             data = top_celldata,
             alpha = 0.5) + 
  geom_line(size = 2, lineend = "round", linejoin= "round") +
  scale_x_log10(labels = scales::label_number(accuracy = 1)) +
  scale_y_log10(labels = scales::label_number(accuracy = 1)) + 
  scale_color_gradient2(low = "grey70", 
                        mid = "forestgreen", 
                        high = "darkgreen", 
                        midpoint = 50) +
  labs(
    x = "Rank", 
    y = "Total UMI count",
    color = "% passing\ncell filter"
  ) + 
  theme(legend.position = c(0, 0),
        legend.justification = c(0,0),
        legend.background = element_rect(color = "grey20", size = 0.25),
        legend.box.margin = margin(rep(5, 4)))
```

The total UMI count of each droplet (barcode) plotted against the rank of that droplet allows visualization of the distribution of sequencing depth across droplets. 
The droplets that are expected to contain cells were identified with [`DropletUtils::emptyDropsCellRanger()`](https://bioconductor.org/packages/release/bioc/html/DropletUtils.html), unless otherwise specified in the `Cell Statistics` table, which uses both the total UMI counts and expressed gene content (adapted from [Lun  _et al._ 2019](https://doi.org/10.1186/s13059-019-1662-y)).
As the boundary between droplets passing and failing this filter is not solely dependent on total UMI count, some regions contain droplets in both categories.
The color in this plot indicates the percentage of droplets in a region passing the filter.

## Cell Read Metrics 

```{r, fig.alt="Total UMI x genes expressed"}
filtered_celldata <- data.frame(colData(filtered_sce)) 

ggplot(filtered_celldata, 
       aes (x = sum,
            y = detected, 
            color = subsets_mito_percent)) +
  geom_point(alpha = 0.3) +
  scale_color_viridis_c(limits = c(0, 100)) + 
  scale_x_continuous(labels = scales::label_number(accuracy = 1)) +
  scale_y_continuous(labels = scales::label_number(accuracy = 1)) +
  labs(x = "Total UMI count",
       y = "Number of genes detected",
       color = "Percent reads\nmitochondrial") + 
  theme(legend.position = c(0, 1),
        legend.justification = c(0,1),
        legend.background = element_rect(color = "grey20", size = 0.25),
        legend.box.margin = margin(rep(5, 4)))
```

The above plot of cell metrics includes only droplets which have passed the `emptyDrops()` filter.
The plot will usually display a strong (but curved) relationship between the total UMI count and the number of genes detected.
Cells with low UMI counts and high mitochondrial percentages may require further filtering.

## miQC Model Diagnostics

```{r, fig.alt="miQC model diagnostics plot", results='asis', warning=FALSE}
if(skip_miQC){
  cat("miQC model not created, skipping miQC plot. Usually this is because mitochondrial gene data was not available.")
}else{
  # remove prob_compromised if it exists, as this will cause errors with plotModel
  filtered_sce$prob_compromised <- NULL
  miQC_model <- metadata(filtered_sce)$miQC_model
  
  if (is.null(miQC_model) || length(miQC_model@components) < 2){
    # model didn't fit, just plot metrics
    miQC_plot <- miQC::plotMetrics(filtered_sce)
  }else{
    miQC_plot <- miQC::plotModel(filtered_sce, model = miQC_model)
  }
  
  miQC_plot + 
    coord_cartesian(ylim = c(0,100)) +
    scale_x_continuous(labels = scales::label_number(accuracy = 1)) +
    labs(x = "Number of genes detected",
         y = "Percent reads mitochondrial") +
    theme(legend.position = c(1, 1),
          legend.justification = c(1,1),
          legend.background = element_rect(color = "grey20", size = 0.25),
          legend.box.margin = margin(rep(5, 4)))
}
```

We calculate the probability that a cell is compromised due to degradation or rupture using [`miQC`](https://bioconductor.org/packages/release/bioc/html/miQC.html) ([Hippen _et al._ 2021](https://doi.org/10.1371/journal.pcbi.1009290)).
This relies on fitting a mixture model using the number of genes expressed by a cell and the percentage of mitochondrial reads.
The expected plot will show a characteristic triangular shape and two model fit lines.
Cells with low numbers of genes expressed may have both low and high mitochondrial percentage, but cells with many genes tend to have a low mitochondrial percentage.
Compromised cells are likely to have a fewer genes detected and higher percentage of mitochondrial reads.

If the model has failed to fit properly, the pattern of cells may differ, and there may not be model fit lines.
This can be the result of a low-quality library or may occur if there is no mitochondrial content, as in the case of a high-quality single-nucleus sample.
In such situations, the calculated probability of compromise may not be valid (see [miQC vignette](https://bioconductor.org/packages/3.13/bioc/vignettes/miQC/inst/doc/miQC.html#when-not-to-use-miqc) for more details).


## Dimensionality Reduction 

```{r message=FALSE}
if(has_processed){
  
  # check that processed SCE contains UMAP and PCA
  if(!"UMAP" %in% reducedDimNames(processed_sce)){
    if(!"PCA" %in% reducedDimNames(processed_sce)){
      
      # if PCA is missing, check for HVG, if no HVG use all genes 
      if(!is.null(processed_meta$highly_variable_genes)){
        subset_genes <- processed_meta$highly_variable_genes
      }else{
        subset_genes <- rownames(processed_sce)
      }
      processed_sce <- scater::runPCA(processed_sce,
                                      subset_row = subset_genes)
    }
    
    processed_sce <- scater::runUMAP(processed_sce, 
                                     dimred = "PCA")
  }
  
  # create UMAP colored by number of genes detected 
  scater::plotUMAP(processed_sce,
                   point_size = 0.3,
                   point_alpha = 0.5,
                   colour_by = "detected") +
    scale_color_viridis_c()+
    theme_bw() + 
    guides(
      color = guide_colorbar(title = "Number of \ngenes detected")
    )
  
}
```

Compromised cells, calculated using `miQC` described above, are filtered and removed from the library prior to downstream analyses. 
The raw counts from all remaining cells are then normalized prior to selection of highly variable genes and dimensionality reduction. 
Here we show the UMAP (Uniform Manifold Approximation and Projection) embeddings for each cell, coloring each cell by the total number of genes detected per cell.

<!-- Next section included only if CITE-seq data is present -->
```{r, child='cite_qc.rmd', eval = has_cite}
```

<!-- Next section only included if multiplex data is present --> 
```{r, child='multiplex_qc.rmd', eval = has_cellhash}

```


# Session Info
<details>
<summary>R session information</summary>
```{r session_info}
if (requireNamespace("sessioninfo", quietly = TRUE)) {
  sessioninfo::session_info()
} else {
  sessionInfo()
}

```
</details>

